---
title: 'Linguistic Data: Quantitative Analysis and Visualisation'
subtitle: 'Lab 3. Student`s t-test. Binomial test. R: Simulating data, boxplots, density plots'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

## Student's t-test

### Aspiration and vowel duration in Icelandic 

This set is based on (Coretta 2017, [link](https://goo.gl/NrfgJm)). This dissertation dealt with the relation between vowel duration and aspiration in consonants. Author carried out a data collection with 5 natives speakers of Icelandic. Then he extracted the duration of vowels followed by aspirated versus non-aspirated consonants. Check out whether vowels before aspirated consonants (like in Icelandic takka ‘key’ [tʰaʰka]) are signiﬁcantly shorter than vowels followed by non-aspirated consonants (like in kagga ‘barrel’ [kʰakka]).  
[Link](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv) to the dataset.

```{r message=FALSE}
Sys.setlocale("LC_ALL", 'en_US.UTF-8')
library(readr)
library(tidyverse)
df <- read_csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv")
```

### Descriptive statistics

A general boxplot:

```{r}
boxplot(df$vowel.dur)
```

Get the number of outliers:

```{r}
length(boxplot(df$vowel.dur)$out)
```

Look at number of observations by groups (aspirated and non-aspirated cases):

```{r}
table(df$aspiration)
```

Choose two subsamples, one for words where vowels are followed by aspirated consonants and another for non-aspirated consonants.

```{r}
asp <- df[df$aspiration == 'yes',]
nasp <- df[df$aspiration == 'no',]
```

Summary for aspirated and non-aspirated cases:

```{r}
summary(asp$vowel.dur)
summary(nasp$vowel.dur)
```

Boxplot by groups:

```{r}
boxplot(df$vowel.dur ~ df$aspiration)
```

More interesting - let us create a boxplot by all groups (see the field `cons1`):

```{r}
boxplot(df$vowel.dur ~ df$cons1)
```

You can compare distribution of `vowel.dur` in asp(irated), fri(cative), nasp(non-aspirated), voi(ced), etc.

We can limit our data to just one type of vowels, say, middle vowels. Therefore, we will work with the same type of a consonant:

```{r}
asp <- df[df$aspiration == 'yes' & df$height == 'mid', ]
nasp <- df[df$aspiration == 'no' & df$height == 'mid', ]
```

Again, here is a summary for a corrected case:

```{r}
summary(asp$vowel.dur)
summary(nasp$vowel.dur)

nrow(asp)
nrow(nasp)
```

### T-test

Let us formulate the null hypothesis, the alternative hypotesis, and apply t-test to our dataset.

```{r}
t.test(asp$vowel.dur, nasp$vowel.dur)
```

By default, R calculates t.test with regard to the bi-directional alternative hypothesis, such as $\mu_1 \neq \mu_2$.

### Unidirectional t-test

H1: $\mu_{asp} \lt \mu_{nasp}$

```{r}
t.test(asp$vowel.dur, nasp$vowel.dur, alternative = "less")
```

### Density plots
```{r, message=FALSE, warning=FALSE}
require(tidyverse)
require(dplyr)
```

Let's get a descriptive summary of our data in a dplyr style.

```{r}
df %>% 
  group_by(aspiration) %>%
  summarise(mean = mean(vowel.dur),
            st.dev = sd(vowel.dur))
```

Density plots can be thought of as plots of smoothed histograms.

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
df %>% 
  ggplot(aes(vowel.dur, fill = aspiration, color = aspiration))+
  geom_density(alpha = 0.4)+
  geom_rug()+
  labs(title = "Vowel duration density plot",
       caption = "Data from (Coretta 2017)",
       x = "vowel duration")
```

Density plot by speaker:

```{r}
df %>% 
  ggplot(aes(vowel.dur, fill = aspiration, color = aspiration))+
  geom_density(alpha = 0.4)+
  geom_rug()+
  facet_wrap(~speaker)+
  labs(title = "Vowel duration density plot, by speaker",
       caption = "Data from (Coretta 2017)",
       x = "vowel duration")
```

and descriptive statistics:

```{r}
df %>% 
  group_by(aspiration, speaker) %>%
  summarise(mean = mean(vowel.dur),
            st.dev = sd(vowel.dur))
```



## Simulating data 

Create a matrix 2 * 3 consisting of 0: 

```{r}
matrix(0, nrow=2, ncol=3)
```

Arrange a vector of 12 values into matrix 3 * 4 arrange by rows:

```{r}
v <- 1:12
m <- matrix(v, n=3, ncol=4, byrow = TRUE)
m
```

Sum of every row in a matrix:

```{r}
rowSums(m)
```

Create a sample of 0 and 1 of size 10.

```{r}
sample(c(0, 1), 10, replace = TRUE)
```

### Sample size and variance 

```{r}
population <- c(1, 2, 3, 4, 5)
sample_size <- 2
means <- rep(NA, 10000)
for (i in 1:10000) {
  samp <- sample(population, size=sample_size, replace=TRUE)
  means[i] <- mean(samp)
}
library(ggplot2)
ggplot() + 
  geom_histogram(aes(means), bins = 100) +
  xlim(c(0, 5))
```

Now you can change the sample size to see how the distribution of means changes. 

See also code for [Simulating variance estimates](https://rstudio-pubs-static.s3.amazonaws.com/237011_27e85fe7e17e4c73a260cb27e60fb012.html)


## Binomial test 

Experiment: toss a coin 10 times and repeat this sequence 10000 times:

```{r}
tosses <- 10
samples <- 10000
dat <- matrix(sample(c(0, 1), tosses * samples, replace=TRUE), ncol=tosses, byrow=TRUE)
```

Calculate `phats` - proportions of heads in each experiment:

```{r}
pbar <- rowSums(dat) / tosses
hist(pbar, breaks=tosses, xlim=c(0, 1))
```

Test $H_0: p = 0.5$:

```{r}
binom.test(3, 10, p=0.5) # 3 out of 10 - a fair coin?
binom.test(2, 10)  # 2 out of 10 - a fair coin?
binom.test(1, 10)  # 1 out of 10 - a fair coin?
```

Load some dataset and check some null hypothesis with binomial test.

```{r}
df <- read.csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv", sep = "\t")
```

Suggest your hypotheses about p of nouns. Look at frequencies:

```{r}
table(df$UPoS)
table(df$UPoS)/sum(table(df$UPoS))
```

Is it enough to make conclusions? No, proceed to formal tests:

```{r}
# select lines with nouns
nouns <- df[df$UPoS=='NOUN',]
total <- nrow(df) # number of trials
nnouns <- nrow(nouns) # number of successes
```

```{r}
# H0: p = 0.6

```

```{r}
# H0: p = 0.4

```

```{r}
# choose lines with one-syllable words at the end




```

```{r}
# you can test on your own for any number of syllables


```

 
#### Supplementary R code

This code generates a dataset that consists of Utterances (strings of letters) and Responces corresponding to each utterance (either 0 or 1)
```{r generating-dataset}
# require(stringi)
n <- 1000 # the number of datapoins
df <- cbind.data.frame(Utterance = stringi::stri_rand_strings(10, 5), # generate a random string
                       Responce = rbinom(n, 1, 0.2))  # generate an answer (either 0 ot 1) randomly with p(1) = 0.2
```

This code run binom.test $n$ times with forward-pipes:
```{r message=FALSE}
require(dplyr)
require(broom)
m <- 5 # sample size in each run
n <- 10 # the number of experiments
dat <- replicate(n=n, expr = sample(0:1, size=m, replace=TRUE)) %>% 
           t() %>%                    # transpose row and columns
           as.data.frame() %>% 
           mutate(ID=row_number(), sum=rowSums(.), m=m) # add ID, row sums, sample size
dat2 <- dat %>% 
            group_by(ID) %>%
            do(tidy(binom.test(.$sum, .$m, alternative = "two.sided"))) %>%
            select(ID, p.value)
```

#### Supplementary materials: Outliers 

By default, boxplot in R is plotted with whiskers and outliers. The usual method to identify outliers (points that lie beyond the extremes of the whiskers) is based on the notion of median, lower (1st) and upper (3rd) quartiles, and certain coefficient like 1.5 below.
Inter-quartile range (IQR) is a difference between the 1st quartile (Q3) and the 3rd quartile (Q1):  
IQR = Q3 - Q1,  
Lower outlier limit = Q1 - 1.5 * IQR,  
Upper outlier limit = Q3 + 1.5 * IQR.  

See also boxplot notches (plotted by request) that extend to Q1 + 1.58 \* IQR/sqrt(n) and Q1 - 1.58 \* IQR/sqrt(n), [figure](https://media.geeksforgeeks.org/wp-content/uploads/fake-data-notch-boxplot.jpg) and are usually interpreted as a 95% confidence interval for the median. 
