---
title: "Homework 1"
output:
  pdf_document: default
  html_document: default
date: 'Deadline: 9 February, 23:59:59'
subtitle: 'Linguistic Data: Quantitative Analysis and Visualisation. Linguistic theory group.'
---

The solutions should be submitted via GitHub.  

## Part 1. A preliminary training  

*Do not use R (RStudio) to solve problems in Part 1.*

### Problem 1

Look at the following histogram and answer the questions.

```{r, echo=FALSE}
set.seed(1234)
x <- rnorm(20, mean=3, sd=6)
hist(x, col="tomato", main="")
```

a. What is the proportion of values in the sample that exceed $5$? Explain your answer.
b. Indicate the interval where the median of this sample can lie. Explain your answer.
c. How the histogram will change if we add an element $7$ to the sample? Explain your answer.

### Problem 2

a. Look at the histograms of two samples. They illustrate the distribution of normalized average reaction time to frequent words (in ms) in two groups of people.

```{r, echo=FALSE, fig.height=3}
set.seed(1234)
sample1 <- rnorm(100, 500, 50)
sample2 <- rnorm(100, 500, 20)
hist(sample1, col="lightblue", xlim=c(100, 900))
```

```{r, echo=FALSE, fig.height=3}
hist(sample2, col="lightgreen", xlim=c(100, 900))
```

Which of the samples has a larger variance? Explain your answer.

\newpage

b. Look at the histograms of two samples.

```{r, echo=FALSE, fig.height=8}
set.seed(12)
sample1 <- rnorm(100, 3, 5)
sample2 <- rnorm(100, 10, 0.8)
par(mfrow=c(2,2))
hist(sample1, col="lightblue", xlim=c(-40, 40))
hist(sample2, col="lightgreen", xlim=c(5, 15))
```

Which of the samples has a larger variance? Explain your answer.

## Part 2

*Do not use R (RStudio) to solve problems in Part 2. Answers for problem 3 will be evaluated. Please paste YES or NO into (empty) code blocks and explain you answer below the block.*

### Problem 3

Below is the histogram of the number of mistakes students made while writing an examination essay in English. Look at the histogram and answer the questions.

```{r, echo=FALSE, fig.height=4}
set.seed(2)
mistakes <- rbeta(1000, 2, 7)*100
hist(mistakes, breaks=60, col= "deepskyblue", ylim=c(0, 40))
```

#### 3.1
Is it true that $50$% students made more than $35$ mistakes?
```

```
Explain your answer below:


#### 3.2
Is it true that most students made no more than $10$ mistakes?
```

```
Explain your answer below:


#### 3.3
Which of the following values is closer to the median of `mistakes`: $10$, $20$, $30$, $40$?
```

```
Explain your answer below:

### Problem 4. Exact binomial test
In a certain language there are two forms of a word "go": normal and dialectical. Whe know that if we select random person from the Country, this person will use normal form with probability 2/3 and dialectical form with probability 1/3. (One person uses only one form all the time.) Researcher suggests that the percentage of people who use dialectical form in a particular City is higher than in the Country. To prove this point, she proceed with the following experiment. Random person from the City is selected and his/her usage of word "go" is recorded. This is repeated $n$ times (the same person theoretically can be chosen more than one time, but the City is large comparatively to $n$, so it rarely occurs in practice).

The results are as following: 20 informants where selected ($n=20$), 17 of them use dialectical form.

Is it enough to say that the percentage of people who use dialectical form in the City is higher than in the Country.

#### 4.1 Hypothesis
State the null hypothesis and alternative.

$H_0$: <YOUR ANSWER HERE>

$H_1$: <YOUR ANSWER HERE>


#### 4.2 Find p-value
Recall that p-value is a probability to get the data that we obtained or "more extreme" (more convincing to reject null hypothesis in favor of alternative) provided that null hypothesis holds. Find p-value for your data.

**Hint.** There is a function `dbinom(k, n, p)` in R that calculates probability to get `k` heads if you toss a coin `n` times and probability to obtain a head in one tossing is `p`. You can use this function here.
```

```

#### 4.3 Conclusion
Will you reject null hypothesis? Use significance level of 5%.


#### 4.4 Answer
Can we claim that we have enough evidence to say that the percentage of people who use dialectical form in the City is higher than in the Country?


## Part 3

*Use R (RStudio) to solve problems in Part 3. Your answers will be evaluated. Please paste R code into R code blocks and explain you answer below the block, if needed.*

### Problem 5 

Here is a sample of respondents' age:  

$44$, $50$, $42$, $64$, $66$, $42$, $72$, $56$, $72$, $54$, $46$, $48$, $48$, $52$, $50$, $66$, $84$.

#### 5.1
Arrange them in a vector and call it `age`. 
```{r}

```

#### 5.2
Examine the type of `age` variable (numeric, character, etc).
```

```

#### 5.3
Plot the histogram of the vector `age` with $5$ bins. Change its color to any you want. (Use either R basic or ggplot2 style for plotting.)
```{r}

```


### Problem 6

Here is a series of words:  
*pie, bar, bar, pie, pie, bar, bar, chart*.

#### 6.1
Arrange elements above in a vector and call it `words` 
```{r}

```

#### 6.2
Calculate the relative frequences of values in `words` measured in percent.
```{r}

```




## Supplementary reading
Use of exact binomial test in linguistic research:  

* Gries, Stefan Th. "Phonological similarity in multi-word units." Cognitive Linguistics 22.3 (2011): 491-510. [Link](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.186.7412&rep=rep1&type=pdf)  
Stefan Gries proves that alliteration is observed in multi-word expressions more often than in general.

* Harald Bayen (2008: 51-52) evaluates the probability of observing exactly one occurrence of the word _hare_ in the corpus sample of 1 mln words given its estimated frequency of 8.23 words per million according to the SELEX frequency database.

On measures of central tendency: 

* Levshina 2015, Chapter 3 (p. 48); Gries 2009, Chapter 1.3 (p. 116). 

