---
title: 'Linguistic Data: Quantitative Analysis and Visualisation'
author: <h4 style="font-style:normal;font-weight:normal">George Moroz, Ilya Schurov, Olga Lyashevskaya</h4>
date: <h4 style="font-weight:normal">25 January 2020</h4>
output:
  html_document: default
---

# `tidyverse`: Loading and transforming data {#tidy_dplyr} 

[_tidyverse_](https://www.tidyverse.org) is a set of packages:

* _ggplot2_, to visualise data 
* _tibble_, to make nice views for dataframes 
* _tidyr_, tidy data format 
* _readr_, to read files in R 
* _purrr_, to do functional programming 
* _dplyr_, to transform data 
* _stringr_, to work with strings 
* _forcats_, to work with categorial variables 

Other useful packages: 

* _readxl_, to read .xls и .xlsx  
* _jsonlite_, to work with JSON  
* _rvest_, to do web scrapping  
* _lubridate_, for time data  
* _tidytext_, to work with texts and corpora  
* _broom_, to transfer statistic models in tidy format  

```{r, message = TRUE}
library("tidyverse")
```

## Loading the data 
### Working directory 

Two functions to set up your working directory: `getwd()` and `setwd(.../path/to/your/directory)`.

### `.csv` files 

`.csv` (comma separated values) is the most popular file format to store the (tabular) data. 
NB [Microsoft Excel нужны некоторые трюки](https://superuser.com/questions/291445/how-do-you-change-default-delimiter-in-the-text-import-in-excel).

### Reading data: `readr`, `readxl` 
* Basic R: `read.csv()`,
* Tydyverse: `read_csv()` (package `readr`) 

```{r, eval = FALSE}
read_csv("...")
```

What's within brackets?

* file name (please check that it exists in your working directory)
```{r, eval = FALSE}
read_csv("my_file.csv")
```

* relative path to the file
```{r, eval = FALSE}
read_csv("data/my_file.csv")
```

* full path to the file
```{r, eval = FALSE}
read_csv("/home/user_name/work/data/my_file.csv")
```

* URL (please check that you are connected to internet)
```{r, eval = FALSE}
read_csv("https://my_host/my_file.csv")
```

Other functions to read `.csv` files:

* `read_tsv()` -- tab-separated files 
* `read_csv2()` -- semicolon-separated files   
* `read_delim(file = "...", delim = "...")` -- any separator set as `delim`

